\chapter{Experiments}
\label{chapter:experiments}

This Chapter describes a series of experiments that were performed in order to
demonstrate the capabilities of the proposed method regarding two areas:
forecasting and interpretability. The forecasting capabilities of the method are
demonstrated by exhibiting how the multi-agent model simulates the prices in
training and testing datasets, and how the system's curve-fitting error
decreased throughout the genetic algorithm's generations. The simulated profits
are also included, for both the training and testing datasets. Regarding the
interpretation capabilities of the method, a description of how the agents
perceive the market is shown. This description also describes what actions the
agents take according to their perceptions.

A total of 135 experiments were performed, but due to space constraints in this
thesis document, only some of the experiments are shown. The reader can find the
plots and interpretations for all the experiments performed in the Git repository of
this thesis here: https://github.com/amherag/PhD-Thesis/.

% All 100 gen, same range of time

\section{Forecasting the Prices of a Financial Market}
\label{section:forecasting-the-prices-of-a-financial-market}



\section{Extracting Insights about a Financial Market}
\label{section:extracting-insights-about-a-financial-market}

\subsection{AUD/USD 4 Agents, 4 Rules, 4 Individuals}
\label{}

{\scriptsize
\begin{itemize}
\item 3 agents — with an average profit of 68 units — perceived in 85\% of the market that a weak resistance above current price — with a hesitancy of 0.147 — is a signal to sell — with a hesitancy of 0.079.
\item 2 agents — with an average profit of 352 units — perceived in 47\% of the market that a moderate resistance below current price — with a hesitancy of 0.134 — is a signal to sell — with a hesitancy of 0.151.
\item 2 agents — with an average profit of 323 units — perceived in 6\% of the market that a strong resistance below current price — with a hesitancy of 0.098 — is a signal to sell — with a hesitancy of 0.024.
\item 2 agents — with an average profit of 323 units — perceived in 55\% of the market that a weak resistance nearby current price — with a hesitancy of 0.04 — is a signal to sell — with a hesitancy of 0.138.
\item 2 agents — with an average profit of 309 units — perceived in 0\% of the market that a strong resistance above current price — with a hesitancy of 0.127 — is a signal to sell — with a hesitancy of 0.14.
\end{itemize}
}